---
title: "Modelling longitudinal data: the individual approach"
author: "Marc Lavielle </br> January 27, 2014 </br> &nbsp; <hr>"

output: 
  html_document:
    toc: true
---
---
<link href="../markdown3.css" rel="stylesheet"></link>  


### Introduction
We will illustrate this modeling process using a pharmacokinetics (PK) example.
The point of this example is to show how several of the tasks described earlier can be easily implemented using the
statistical software \R. Some knowledge of statistics and \R is necessary for following the various modeling steps and implementing the methods.

### The data
Let us consider a dose D = 50mg of a drug administered orally to a patient at time $t=0$. The concentration of the drug in the bloodstream $(y_j, 1\leq j \leq 12)$ is then measured at times $(t_j, 1\leq j \leq 12)$:


 First, we import the data stored in the data file `individualFitting_data.txt` and plot it to have an initial look:

```{r , fig.width=5}
pk1=read.table("individualFitting_data.txt",header=T)
pk1

t=pk1$time
y=pk1$concentration
n=length(t)
plot(t,y,xlab="time (hours)",ylab="concentration (mg/l)")
abline(a=0,b=0,lty=3,col="grey")
```

### Fitting two PK models
$$
\def\ka{k_a}
\def\kel{k}
\def\teps{\varepsilon}
$$

We are going to consider two possible structural models that may describe
the observed time-course of the concentration:

* A one-compartment model with first-order absorption and first-order elimination:
$$
\begin{aligned}
\phi_1 &= (\ka, V, \kel) \\
f_1(t ; \phi_1) &= \frac{D\, \ka}{V(\ka-\kel)} \left( e^{-\kel \, t} - e^{-\ka \, t} \right).
\end{aligned}
$$
```{r}
predc1=function(t,x){
   f=50*x[1]/x[2]/(x[1]-x[3])*(exp(-x[3]*t)-exp(-x[1]*t)) 
return(f)}
```

* A one-compartment model with zero-order absorption and first-order elimination:
$$
\begin{aligned}
\phi_2 &= (T_{k0}, V, \kel) \\
f_2(t ; \phi_2) &= \left\{  \begin{array}{ll}
                         \frac{D}{V \,T_{k0} \, \kel} \left( 1- e^{-\kel \, t} \right) & {\rm if }\ t\leq T_{k0} \\[0.3cm]
                         \frac{D}{V \,T_{k0} \, \kel} \left( 1- e^{-\kel \, T_{k0}} \right)e^{-\kel \, (t- T_{k0})} & {\rm otherwise} .
                       \end{array}
\right.
\end{aligned}
$$
```{r}
predc2=function(t,x){ 
   f=50/x[1]/x[2]/x[3]*(1-exp(-x[3]*t))
   f[t>x[1]]=50/x[1]/x[2]/x[3]*(1-exp(-x[3]*x[1]))*
     exp(-x[3]*(t[t>x[1]]-x[1]))
return(f)}
```
We then define two models ${\cal M}_1$ and ${\cal M}_2$ that assume (for now)  constant residual error models:
$$
\begin{aligned}
{\cal M}_1  : \quad y_j & =  f_1(t_j ; \phi_1) + a_1\teps_j \\
{\cal M}_2  : \quad y_j & =  f_2(t_j ; \phi_2) + a_2\teps_j .
\end{aligned}
$$

We can fit these two models to our data by computing the maximum likelihood estimate (MLE) $\hat{\psi}_1=(\hat{\phi}_1,\hat{a}_1)$ and $\hat{\psi}_2=(\hat{\phi}_2,\hat{a}_2)$ of $\psi$  under each model. We can use the `nlm` function for nonlinear minimization. This function requires an initial value $\psi_0 = (\phi_0, a_0)$:

```{r}
fmin1=function(x,y,t){
   f=predc1(t,x)
   g=x[4]
   e=sum( ((y-f)/g)^2 + log(g^2))
return(e)}

fmin2=function(x,y,t){
   f=predc2(t,x)
   g=x[4]
   e=sum( ((y-f)/g)^2 + log(g^2))
return(e)}

options(warn=-1)
pk.nlm1=nlm(fmin1, c(0.3,6,0.2,1), y, t, hessian="true")
psi1=pk.nlm1$estimate
pk.nlm2=nlm(fmin2, c(3,10,0.2,4), y, t, hessian="true")
psi2=pk.nlm2$estimate

cat(" psi1 =",psi1,"\n","psi2 =",psi2)
```
### Evaluating and selecting PK models
The estimated parameters $\hat{\phi}_1$ and $\hat{\phi}_2$ can then be used for computing the predicted concentrations $\hat{f}_1(t)$ and $\hat{f}_2(t)$ under both models at any time $t$. These curves can  be plotted over the original data and compared:


```{r}
tc=seq(from=0,to=25,by=0.1)
phi1=psi1[c(1,2,3)]
fc1=predc1(tc,psi1)
phi2=psi2[c(1,2,3)]
fc2=predc2(tc,psi2)

plot(t,y,ylim=c(0,4.1),xlab="time (hours)",ylab="concentration (mg/l)")
lines(tc,fc1, type = "l", col = "black", lwd=1, lty=1)
lines(tc,fc2, type = "l", col = "black", lwd=1, lty=4)
abline(a=0,b=0,lty=3,col="grey")
legend(13,4,
       c("observations","first-order absorption","zero-order absorption"), 
       lty=c(-1,1,4), pch=c(1,-1,-1), lwd=1) 
```

We clearly see that a much better fit is obtained with model ${\cal M}_2$, i.e.,  assuming a zero-order absorption process.

Another useful goodness-of-fit plot can be obtained by plotting the observations $(y_j)$ versus the predictions $\hat{y}_j=f(t_j ; \hat{\psi})$ given by the models:

```{r}
f1=predc1(t,phi1)
f2=predc2(t,phi2)

par(mfrow= c(1,2))
plot(f1,y,xlim=c(0,4),ylim=c(0,4), main="model 1")
abline(a=0,b=1,lty=1)
plot(f2,y,xlim=c(0,4),ylim=c(0,4), main="model 2")
abline(a=0,b=1,lty=1)
```

### Model selection
According to the previous diagnostic plots, ${\cal M}_2$ would seem to have a slight edge.
This can be tested more analytically using the BIC:

```{r}
deviance1=pk.nlm1$minimum + n*log(2*pi)
bic1=deviance1+log(n)*length(psi1)
deviance2=pk.nlm2$minimum + n*log(2*pi)
bic2=deviance2+log(n)*length(psi2)
cat(" bic1 =",bic1,"\n","bic2 =",bic2)
```

A smaller BIC is better. Therefore, this result also suggests that model ${\cal M}_2$ should be selected.

### Fitting different error models
For the moment, we have only considered  constant error models. However, the *observations vs predictions* plot  hints that the amplitude of the residual errors may increase with the size of the predicted value. Let us therefore take a closer look at four different residual error models, each of which we will associate with the *best* structural model $f_2$:

${\cal M}_2$  constant error model:  $y_j=f_2(t_j;\phi_2)+a_2\teps_j$

${\cal M}_3$  proportional error model:  $y_j=f_2(t_j;\phi_3)+b_3f_2(t_j;\phi_3)\teps_j$

${\cal M}_4$ combined error model:  $y_j=f_2(t_j;\phi_4)+(a_4+b_4f_2(t_j;\phi_4))\teps_j$ 

${\cal M}_5$  exponential error model: $\log(y_j)=\log(f_2(t_j;\phi_5)) + a_5\teps_j$.


The three new ones need to be entered into R:

```{r}
fmin3=function(x,y,t){
   f=predc2(t,x)
   g=x[4]*f
   e=sum( ((y-f)/g)^2 + log(g^2))
return(e)}

fmin4=function(x,y,t){
   f=predc2(t,x)
   g=abs(x[4])+abs(x[5])*f
   e=sum( ((y-f)/g)^2 + log(g^2))
return(e)}

fmin5=function(x,y,t){
   f=predc2(t,x)
   g=x[4]
   e=sum( ((log(y)-log(f))/g)^2 + log(g^2))
return(e)}
```

We can now compute the MLEs $\hat{\psi}_3=(\hat{\phi}_3,\hat{b}_3)$, $\hat{\psi}_4=(\hat{\phi}_4,\hat{a}_4,\hat{b}_4)$ and $\hat{\psi}_5=(\hat{\phi}_5,\hat{a}_5)$ of $\psi$  under models ${\cal M}_3$, ${\cal M}_4$  and ${\cal M}_5$:


```{r}
pk.nlm3=nlm(fmin3, c(phi2,0.1), y, t, hessian="true")
psi3=pk.nlm3$estimate

pk.nlm4=nlm(fmin4, c(phi2,1,0.1), y, t, hessian="true")
psi4=pk.nlm4$estimate
psi4[c(4,5)]=abs(psi4[c(4,5)])
 
pk.nlm5=nlm(fmin5, c(phi2,0.1), y, t, hessian="true")
psi5=pk.nlm5$estimate

cat(" psi3 =",psi3,"\n","psi4 =",psi4,"\n","psi5 =",psi5)
```


### Selecting the error model
As before, these curves can be plotted over the original data and compared:

```{r}
phi3=psi3[c(1,2,3)]
fc3=predc2(tc,phi3)
phi4=psi4[c(1,2,3)]
fc4=predc2(tc,phi4)
phi5=psi5[c(1,2,3)]
fc5=predc2(tc,phi5)

par(mfrow= c(1,1))
plot(t,y,ylim=c(0,4.1),xlab="time (hours)",ylab="concentration (mg/l)")
lines(tc,fc2, type = "l", lwd=1)
lines(tc,fc3, type = "l", lwd=1, lty=3)
lines(tc,fc4, type = "l", lwd=1, lty=4)
lines(tc,fc5, type = "l", lwd=1, lty=2)
abline(a=0,b=0,lty=3,col="grey")
legend(13,4,c("observations","constant error model",
      "proportional error model", "combined error model",
      "exponential error model"), lty=c(-1,1,3,4,2), pch=c(1,-1,-1,-1,-1)) 

f2=predc2(t,phi2)
f3=predc2(t,phi3)
f4=predc2(t,phi4)
f5=predc2(t,phi5)

par(mfrow= c(2,2))
plot(f2,y,xlim=c(0,4),ylim=c(0,4), main="model 2")
abline(a=0,b=1,lty=1)
plot(f3,y,xlim=c(0,4),ylim=c(0,4), main="model 3")
abline(a=0,b=1,lty=1)
plot(f4,y,xlim=c(0,4),ylim=c(0,4), main="model 4")
abline(a=0,b=1,lty=1)
plot(f5,y,xlim=c(0,4),ylim=c(0,4), main="model 5")
abline(a=0,b=1,lty=1)
```

The  predicted concentrations obtained with models ${\cal M}_3$, ${\cal M}_4$  and ${\cal M}_5$ are quite similar.
We now calculate the BIC for each:


```{r}
deviance3=pk.nlm3$minimum + n*log(2*pi)
bic3=deviance3 + log(n)*length(psi3)
deviance4=pk.nlm4$minimum + n*log(2*pi)
bic4=deviance4 + log(n)*length(psi4)
deviance5=pk.nlm5$minimum + 2*sum(log(y)) + n*log(2*pi)
bic5=deviance5 + log(n)*length(psi5)

cat(" bic3 =",bic3,"\n","bic4 =",bic4,"\n","bic5 =",bic5)
```

All of these BIC are lower than the constant residual error one.

There is not however a large difference between them, though the proportional and combined error models give the smallest and essentially identical BIC.  We will use the combined error model ${\cal M}_4$ in the following (the same types of analyses could be done with the proportional error model).

A 90\% confidence interval for $\psi_4$ can derived from the Hessian (i.e., the square matrix of second-order partial derivatives)  of the objective function (i.e., -2 $\times \ LL$):

```{r}
alpha=0.9
df=n-length(phi4)
I4=pk.nlm4$hessian/2
H4=solve(I4)
s4=sqrt(diag(H4)*n/df)
delta4=s4*qt(0.5+alpha/2,df)
ci4=matrix(c(psi4-delta4,psi4+delta4),ncol=2)
print(ci4)
```

We can also calculate a $90\%$ confidence interval for $f_4(t)$ using the central limit theorem:

```{r}
nlpredci=function(phi,f,H){
   dphi=length(phi)
   nf=length(f)
   H=H*n/(n-dphi)
   S=H[seq(1,dphi),seq(1,dphi)]
   G=matrix(nrow=nf,ncol=dphi)
   for (k in seq(1,dphi)) {
      dk=phi[k]*(1e-5)
      phid=phi 
      phid[k]=phi[k] + dk
      fd=predc2(tc,phid)
      G[,k]=(f-fd)/dk 
   }
   M=rowSums((G%*%S)*G)
   deltaf=sqrt(M)*qt(0.5+alpha/2,df)
return(deltaf)}

deltafc4=nlpredci(phi4,fc4,H4)
```

This can then be plotted:

```{r}
par(mfrow= c(1,1))
plot(t,y,ylim=c(0,4.5),xlab="time (hours)",ylab="concentration (mg/l)")
lines(tc,fc4, type = "l", lwd=2)
lines(tc,fc4-deltafc4, type = "l", lwd=1, lty=3)
lines(tc,fc4+deltafc4, type = "l", lwd=1, lty=3)
abline(a=0,b=0,lty=3,col="grey")
legend(10.5,4.5,c("observed concentrations","predicted concentration",
  "PI for predicted concentration"),lty=c(-1,1,3), pch=c(1,-1,-1), lwd=c(2,2,1)) 
```

Alternatively, prediction intervals for $\hat{\psi}_4$, $\hat{f}_4(t;\hat{\psi}_4)$ and new observations at any time $t$ can be estimated by Monte Carlo simulation:

```{r}
f=predc2(t,phi4)
a4=psi4[4]
b4=psi4[5]
g=a4+b4*f
dpsi=length(psi4)
nc=length(tc)
N=500
qalpha=c(0.5 - alpha/2,0.5 + alpha/2)
PSI=matrix(nrow=N,ncol=dpsi)
FC=matrix(nrow=N,ncol=nc)
Y=matrix(nrow=N,ncol=nc)
for (k in seq(1,N)) {
   eps=rnorm(n)
   ys=f+g*eps
   pk.nlm=nlm(fmin4, psi4, ys, t)
   psie=pk.nlm$estimate
   psie[c(4,5)]=abs(psie[c(4,5)])
   PSI[k,]=psie
   fce=predc2(tc,psie[c(1,2,3)])
   FC[k,]=fce
   gce=a4+b4*fce
   Y[k,]=fce + gce*rnorm(1)
}

mean4s=apply(PSI,2,mean)
median4s=apply(PSI,2,median)
sd4s=apply(PSI,2,sd)
ci4s=matrix(nrow=dpsi,ncol=2)
for (k in seq(1,dpsi)){
   ci4s[k,]=quantile(PSI[,k],qalpha,names=FALSE)
}
print(matrix(c(mean4s,median4s,sd4s),nrow=dpsi))
print(ci4s)

cifc4s=matrix(nrow=nc,ncol=2)
ciy4s=matrix(nrow=nc,ncol=2)
for (k in seq(1,nc)){
   cifc4s[k,]=quantile(FC[,k],qalpha,names=FALSE)
   ciy4s[k,]=quantile(Y[,k],qalpha,names=FALSE)
}

par(mfrow= c(1,1))
plot(t,y,ylim=c(0,4.5),xlab="time (hours)",ylab="concentration (mg/l)")
lines(tc,fc4, type = "l", lwd=2)
lines(tc,cifc4s[,1], type = "l", lwd=1, lty=3)
lines(tc,cifc4s[,2], type = "l", lwd=1, lty=3)
lines(tc,ciy4s[,1], type = "l",  lwd=1, lty=2)
lines(tc,ciy4s[,2], type = "l",  lwd=1, lty=2)
abline(a=0,b=0,lty=3,col="grey")
legend(10.5,4.5,c("observed concentrations","predicted concentration",
       "PI for predicted concentration","PI for observed concentrations"),
       lty=c(-1,1,3,2), pch=c(1,-1,-1,-1), lwd=c(2,2,1,1)) 
```

